Bulk Semantic Annotation with a Partially Known Knowledge Base

### 2024-12-11 doc/cicm25-sysdesc/sysdesc.tex
more

\begin{newpart}{MK: looking at the problem in general and compare it to the informal
    authoring and formal authoring problems e.g. programming in an IDE.}
We contend that semantic authoring is a problem that is distinct from traditional
(informal) authoring and authoring fully formal corpora (e.g. programs or formalizations),
while sharing some of the aspects of either.

@@

In fully formal authoring e.g. for programming, the content commons is given formally in
the form of the libraries of the respective programming language, correspondingly the
authored content is also formal -- a program -- and intended for processing by a computer,
for which the formality is a prerequisite. Formal authoring is usually supported by an
integraded development environment (IDE) that makes use of the fully formal content and
offers services like semantic tab completion

\ednote{talk about formalization in logic -- e.g. as theorem prover libraries as well}
\end{newpart}

@@

The \snify\ednote{explain name? write as sn-ify?}
system (see \cite{stextools:git}) is a simple command-line tool
that creates a database of symbol-verbalization pairs, analyzes \sTeX source files,
and then steps through all word occurrences in the document that match a
verbalization in the database.
Matching is based on word stems using off-the-shelf stemmers, which allows us to
e.g.\ match ``utilities'' with ``utility''.
For each matched word, the user is presented with an annotation choice and
interactions that allow to fine-tune the wordwise annotation workflow. \Cref{fig:snify}
shows a typical situation.

>>>>

We contend that semantic authoring is a problem that is distinct from traditional
(informal) authoring and authoring fully formal corpora (e.g. programs or formalizations),
while sharing some of the aspects of either.

@@

In fully formal authoring e.g. for programming, the content commons is given formally in
the form of the libraries of the respective programming language, correspondingly the
authored content is also formal -- a program -- and intended for processing by a computer,
for which the formality is a prerequisite. Formal authoring is usually supported by an
integraded development environment (IDE) that makes use of the fully formal content and
offers services like semantic highlighting, tab completion, jumping to variable
declarations, function definitions, and interface definitions, as well as refactoring
support. All of these rely on a syntactic/semantic analysis of the formal content. The
situation is very similar for authoring of formal knowledge, e.g. in program verification
of the formalization of mathematics.

In semantic authoring, e.g. the \sTeX sources of educational content for the \ALeA system
presented above. The base of the text is classical narrative text to which most of the
constraints of classical authoring apply. The exception is the authoring-time tailoring to
the competencies of the assumed reader, which can (partially) be automated by systems like
\ALeA. Note that it is not a restriction that \ALeA is initially targeted to tertiary
education; the underlying mechanisms generalize to all scientific communication and
technical documents. After all, the purpose is to transfer information (instruct), even if
the roles of instructor and instructee are not predetermined by academic seniority.

But in addition to the base text, semantic authoring also has to provide the semantic
annotations that drive the personalization and interaction. Techniques from formal
authoring could help here, but \ednote{cite Luka's CICM paper \cite{VreWelKam:tsmmdui24}
  for the formulae; introduce the \sTeX way of writing content formulae above.}

@@

The \snify utility (from ``\texttt{\textbackslash sn}-ify'' for adding term references;
see \cite{stextools:git}) is a simple command-line tool that creates a database of
symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
occurrences in the document that match a verbalization in the database.  Matching is based
on word stems using off-the-shelf stemmers, which allows us to e.g.\ match ``utilities''
with ``utility''.  For each matched word, the user is presented with an annotation choice
and interactions that allow to fine-tune the wordwise annotation
workflow. \Cref{fig:snify} shows a typical situation.
